A tool set for random number generation on GPUs in R

### 2021-11-06 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
modify rexp kernel
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2948 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\Address{
  Ruoyong Xu\\
  Department of Statistics\\
  University of Toronto\\
  Universit\"at Innsbruck\\
  Universit\"atsstr.~15\\
  6020 Innsbruck, Austria\\
  E-mail: \email{ruoyong.xu@mail.utoronto.ca}}

\Address{
  Achim Zeileis\\
  Journal of Statistical Software\\
  \emph{and}\\
  Department of Statistics\\
  University of Toronto\\
  Universit\"at Innsbruck\\
  Universit\"atsstr.~15\\
  6020 Innsbruck, Austria\\
  E-mail: \email{ruoyong.xu@mail.utoronto.ca}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
%\SweaveOpts{concordance=FALSE}

@@

The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host using the MRG31k3p RNG. If we want to use the streams in work-items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work-item copies the current states only of the streams to its private memory. \pkg{gpuRandom} is able to create streams both on the host and on the GPU device. \fct{createStreamsCpu} is an interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on host as \proglang{R} matrices, see the \proglang{R} output below, where we created 4 streams on the host using \fct{createStreamsCpu}, after transposing \code{myStreamsCpu}, each column of the matrix is a stream, creating streams on host makes it easy for us to view and check the streams behavior. Substreams are not created in \pkg{gpuRandom} because currently we use one stream per work-item. 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# creating streams on CPU
library("gpuR")
library("gpuRandom")
myStreamsCpu <- createStreamsCpu(n=4, initial=12345)
t(myStreamsCpu)
##               [,1]       [,2]       [,3]       [,4]
## current.g1.1 12345  336690377  502033783  739421137
## current.g1.2 12345  597094797 1322587635 1475938232
## current.g1.3 12345 1245771585 1964121530  730262207
## current.g2.1 12345   85196284 1949818481 1630192198
## current.g2.2 12345  523477687 1607232546  324551134
## current.g2.3 12345 2094976052 1462898381  795289868
## initial.g1.1 12345  336690377  502033783  739421137
## initial.g1.2 12345  597094797 1322587635 1475938232
## initial.g1.3 12345 1245771585 1964121530  730262207
## initial.g2.1 12345   85196284 1949818481 1630192198
## initial.g2.2 12345  523477687 1607232546  324551134
## initial.g2.3 12345 2094976052 1462898381  795289868
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{createStreamsCpu} has two arguments:
\begin{itemize}
\itemsep0em 
  \item \code{n}: number of streams to create.
  \item \code{seed}: vector of length 6, recycled if shorter. Default is set to a vector of six ``12345''.
\end{itemize}
\fct{createStreamsGpu} is adapted from the clrngMrg31k3pCreateStreams() and creates streams directly on a GPU device. We can have streams on GPU by either converting the \code{myStreamsCpu} to a `\code{vclMatrix}' or producing streams using \fct{createStreamsGpu}. Setting a seed same as the MRG31k3p default seed in \fct{createStreamsGpu}, we can produce on device exactly the same streams as those created on host as follows,
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# creating streams on GPU
myStreamsGpu1 = vclMatrix(myStreamsCpu)
myStreamsGpu2 = createStreamsGpu(n=4, initial=12345)
range(as.matrix(myStreamsGpu1 - myStreamsGpu2))
## [1] 0 0
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{createStreamsGpu} also has the two arguments \code{n} and \code{seed}, \code{seed} is by default a vector of six ``12345''. \fct{createStreamsGpu} is similar to \code{.Random.seed} in \proglang{R}. By setting the RNG status with the \proglang{R} function \fct{set.seed}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \fct{createStreamsGpu} and \fct{createStreamsCpu}, with the \code{seed} argument, we make the sequence of streams repeatable.

\subsection{Using streams to generate uniform random numbers}
The streams created sequentially on a GPU are then used in work-items that executes in parallel on a GPU device. In \pkg{gpuRandom} each work-item takes one distinct stream to generate random numbers, so the number of streams should always be equal to the number of total work-items in use. The main part of the kernel (\proglang{OpenCL} functions for execution on the device) for generating uniform random numbers is shown in Listing \ref{lst:uniformkernel}. Kernels are written in the \proglang{OpenCL C} language, in which \code{__kernel} declares a function as a kernel, pointer kernel arguments must be declared with an address space qualifier, for example \code{__global} and \code{__local}. Here the pointers to \code{streams} and to the output matrix \code{out} on the global memory are passed to the kernel as arguments. \code{Nrow} and \code{Ncol} represent row and column number of the matrix \code{out} respectively. \code{NpadCol} is the internal number of columns of \code{out}, \code{NpadStreams} is the internal number of columns of the \code{streams} (these variables are defined in macros not shown here). Each stream's current state is copied to the private memory of each work-item by the function \fct{streamsToPrivate}. \code{index} represents the position of the work-item when mapping it to a one-dimensional data structure. The function \fct{clrngMrg31k3pNextState} generates an uniform random integer called \code{temp} between 1 and 2147483647,  the value of \code{temp} is then scaled to be in the interval $(0,1)$ by multiplying it by a constant \code{mrg31k3p\_NORM\_cl}, which is defined to be $1/2147483648$, if to generate double-precision random numbers. (ADD THINGS ABOUT RANDOM INTEGERS) At the end of generating random numbers, streams are transferred back to global memory through the function \fct{streamsFromPrivate}.
\begin{framed}
\begin{lstlisting}[language=C,basicstyle=\small,label={lst:uniformkernel}]
 __kernel void mrg31k3pMatrix(
    __global int* streams,
    __global double* out){

@@

    }//Dcol
 }//Drow
streamsFromPrivate(streams,g1,g2,startvalue);
}
\end{lstlisting}
\end{framed}
This kernel executes over a 2-dimensional NDRange. Figure \ref{fig:1} illustrates how this kernel executes in practice with a simple example. Suppose we have a $2 \times 4$ NDRange ((a) in Figure \ref{fig:1}) and want to create a $4 \times 6$ matrix ((b) in Figure \ref{fig:1}) of uniform random numbers. In each grid of (a) that represents a work-item, there are two rows of numbers, the number in the above row is the stream index (from 0 to 7), and a pair of numbers in the second row is the 2-dimensional global ID of the work-item. Each stream object is passed to each work-item. The number in a cell of the matrix corresponds to the stream index, it indicates which stream generates a random number for this matrix cell. For example, stream 0 which is allocated in work-item $(0,0)$ generates a random number in matrix $(0,0)$ in the first iteration, then moves on to matrix $(0,2)$ and generates a random number there in the second iteration, in the third iteration it generates a random number in matrix $(0,4)$ and so on until it goes out of the matrix range in the next iteration.
\begin{figure}[H]%
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f1} }}%
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f2} }}%
    \caption{name?\label{fig:1}}%
\end{figure}

Now we use the streams created in section \ref{create_streams} to generate two vectors of double-precision i.i.d. $U(0,1)$ random numbers with the function \fct{runif}. To view the generated random numbers, we need to convert them to R vectors or matrices, by doing this, the random numbers are moved from the global memory of the GPU device to the host.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
as.vector(gpuRandom::runif(n = 6, streams = myStreamsGpu1, Nglobal = c(1,
    nrow(myStreamsGpu1)), type = "double"))
## [1] 0.7353245 0.5180770 0.6142074 0.2319392 0.1100781 0.3619766
as.vector(gpuRandom::runif(n = 10, streams = myStreamsGpu2, Nglobal = c(1,
    nrow(myStreamsGpu1)), type = "double"))[-(1:6)]
## [1] 0.6487742 0.1112075 0.3661944 0.5018562
\end{verbatim}
\end{kframe}
\end{knitrout}
The arguments of the \fct{gpuRandom::runif} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams for random number generation.
  \item \code{Nglobal}: global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' or ``integer'' format of generated random numbers.
\end{itemize}
\fct{runif} returns the same sequence of random numbers if it uses the same streams (here \code{myStreamsGpu1} and \code{myStreamsGpu2}), not streams starting from the same states. Reusing \code{myStreamsGpu1} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream advances by one position. %Below shows the \code{myStreamsGpu1} after it being used.

@@

\subsection{Plot the uniform random numbers}
We evaluate the distribution of 50000 GPU-generated double-precision uniform random numbers by making a density histogram  (the left panel in Figure \ref{fig0}), and plotting its empirical cumulative distribution and overlaying the theoretical cumulative distribution (the right panel in Figure \ref{fig0}). The two cumulative distribution lines seem to overlap completely, suggesting the data generated by \fct{runif} does come from a standard uniform distribution.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n = 512)
random_u<-as.vector(runif(n=50000, streams=streams, Nglobal=c(64,8), type="double"))
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{figure}[H]
\centering

\caption{Histogram and empirical cumulative distribution of the GPU-generated random numbers.\label{fig0}}
\end{figure}

\section{Normal random number generation}
We apply the Box-Muller transformation to $U(0,1)$ random numbers to generate standard Gaussian random numbers. As shown in Algorithm \ref{algorithm1}, Box-Muller algorithm takes two independent, standard uniform random variables $U_1$ and $U_2$ and produces two independent, standard Gaussian random variables $X$ and $Y$, where $R$ and $\Theta$ are polar coordinate random variables. %The derivation basically uses transformation from Cartesian coordinates to polar coordinates to represent two independent, Normally distributed random variables. 
The Box-Muller algorithm turns out to be the best choice for Gaussian transform on GPU compared to other transform methods \citep{howes2007efficient}, because this algorithm has no branching or looping, which are the things GPU is not good at. %but GPU is good at the high computational load of sine and cosine functions in the algorithm.

@@

\begin{figure}[H]%
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f3} }}%
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f4} }}%
    \caption{name???}%
    \label{fig3}
\end{figure}

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{gpuRandom::rnorm}. Both process are not time-consuming, however, using \fct{gpuRandom::rnorm} with $512 \times 128$ work-items is around 120 times (the number may fluctuate a bit from executions) faster than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n = 512 * 128)
system.time(gpuRandom::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.133   0.000   0.133
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   6.608   0.796   7.399
\end{verbatim}
\end{kframe}
\end{knitrout}

\section{Monte Carlo simulation for Fisher's exact test}
One application of GPU-generated random numbers in \pkg{gpuRandom} is Monte Carlo simulation for Fisher's exact test. Fisher’s exact test is applied for analyzing usually $2 \times 2$ contingency tables when one of the expected values in table is less than 5. Different from methods which rely on approximation, Fisher's exact test computes directly the probability of obtaining each possible combination of the data for the same row and column totals (marginal totals) as the observed table, and get the exact p-value by adding together all the probabilities of tables as extreme or more extreme than the one observed. However, when the observed table gets large in terms of sample size and table dimensions, the number of combinations of cell frequencies with the same marginal totals gets very large, \cite[][p. 23]{mehta2011ibm} shows a $5 \times 6$ observed table that has 1.6 billion possible tables. Calculating the exact P-values may lead to very long run-time and can sometimes exceed the memory limits of your computer. Hence, the option \code{simulate.p.value = TRUE} in \fct{stats::fisher.test} is provided, which enables computing p-values by Monte Carlo simulation for tables larger than 2 $\times$ 2.

@@

The arguments of the \fct{gpuRandom::fisher.sim} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{x}: a vclMatrix.
  \item \code{N}: number of simulation runs.
  \item \code{streams}: streams.
  \item \code{type}: ``double'' or ``float'' or returned test statistics.
  \item \code{returnStatistics}: if TRUE, return all test statistics. 
  \item \code{Nglobal}: global index space. Default is set as $(64,16)$.
\end{itemize}
We show the advantage of \fct{gpuRandom::fisher.sim} by computing the p-values for two real data examples: one with a relatively big p-value and another with a very small p-value, and compare the run-time with using \fct{stats::fisher.test} for each of the data sets on two computers: one with a very good CPU and an ordinary GPU, the other is equipped with an excellent GPU and an ordinary CPU. The \proglang{R} outputs for testing on computer 2 is shown in the following.

@@

\subsection{Comparing run-time: Month data example}

@@

month_gpu<-vclMatrix(month,type="integer")
result_month <- fisher.sim(month_gpu, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(256,128))
result_month$simNum
## [1] 1015808
result_month$counts
## [1] 410533
result_month$p.value
## [1] 0.4041449
\end{verbatim}
\end{kframe}
\end{knitrout}
We requested one million simulations on GPU while the actual number of simulation is 1015808, and got 403274 cases whose test statitsic is below the observed threshold, so the p-value from \fct{gpuRandom::fisher.sim} for this table is about 0.40375, which is close enough to the p-value from \fct{stats::fisher.test}: 0.40380.  
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(fisher.sim(month_gpu, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = c(256,128)))
##    user  system elapsed 
##   0.702   0.005   0.707
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{gpuRandom::fisher.sim} takes about 0.32 seconds, \pkg{gpuRandom} accounts for 2.2\% of the elapsed time on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
stats::fisher.test(month,simulate.p.value = TRUE,B=1015808)$p.value
## [1] 0.4029291
system.time(stats::fisher.test(month,simulate.p.value = TRUE,B=1015808))
##    user  system elapsed 
##  14.762   0.003  14.757
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{stats::fisher.test} takes about 9.8 seconds (time varies a little bit from executions) for 1015808 simulations, has a p-value of 0.4035.

\subsection{Comparing run-time: Week data example}

>>>>

\Address{
  Ruoyong Xu, Patrick Brown\\
  Department of Statistics\\
  University of Toronto\\
  700 University Ave., Toronto, ON M5G 1Z5, Canada\\
  E-mail: \email{ruoyong.xu@mail.utoronto.ca}, \email{patrick.brown@utoronto.ca}}

\Address{
  Pierre L’Ecuyer\\
  Department of Computer Science and Operations Research\\
  University of Montr\'eal\\
  Pavillon André-Aisenstadt 2920, chemin de la Tour Montréal QCH3T 1J4\\
  E-mail: \email{lecuyer@iro.umontreal.ca}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
%\SweaveOpts{concordance=FALSE}

@@

The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host using the MRG31k3p RNG. If we want to use the streams in work-items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work-item copies the current states only of the streams to its private memory. \pkg{gpuRandom} is able to create streams both on the host and on the GPU device. \fct{createStreamsCpu} is an interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on host as \proglang{R} matrices, see the \proglang{R} output below, where we created 4 streams on the host using \fct{createStreamsCpu}, after transposing \code{myStreamsCpu}, each column of the matrix is a stream, creating streams on host makes it easy for us to view and check the streams behavior. Substreams are not created in \pkg{gpuRandom} because currently we use one stream per work-item. 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# creating streams on CPU
library("gpuR")
library("gpuRandom")
myStreamsCpu <- createStreamsCpu(n=4, initial=12345)
t(myStreamsCpu)
##               [,1]       [,2]       [,3]       [,4]
## current.g1.1 12345  336690377  502033783  739421137
## current.g1.2 12345  597094797 1322587635 1475938232
## current.g1.3 12345 1245771585 1964121530  730262207
## current.g2.1 12345   85196284 1949818481 1630192198
## current.g2.2 12345  523477687 1607232546  324551134
## current.g2.3 12345 2094976052 1462898381  795289868
## initial.g1.1 12345  336690377  502033783  739421137
## initial.g1.2 12345  597094797 1322587635 1475938232
## initial.g1.3 12345 1245771585 1964121530  730262207
## initial.g2.1 12345   85196284 1949818481 1630192198
## initial.g2.2 12345  523477687 1607232546  324551134
## initial.g2.3 12345 2094976052 1462898381  795289868
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{createStreamsCpu} has two arguments:
\begin{itemize}
\itemsep0em 
  \item \code{n}: number of streams to create.
  \item \code{initial}: vector of length 6, recycled if shorter.
\end{itemize}
We can have streams on GPU by either converting the \code{myStreamsCpu} to a `\code{vclMatrix}' or producing streams using \fct{createStreamsGpu}. \fct{createStreamsGpu} is adapted from the clrngMrg31k3pCreateStreams() and creates streams directly on a GPU device. Setting a seed same as the MRG31k3p default seed in \fct{createStreamsGpu}, we produce on device exactly the same streams as those created on host as follows. Notice that the code is just for showing \fct{createStreamsGpu} and \fct{createStreamsCpu} are \proglang{R} interface functions for \fct{clrngMrg31k3pCreateStreams} on GPU and CPU respectively, in practice it is not recommended to create a new stream that has the same initial seed as another existing stream in one simulation program.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# creating streams on GPU
myStreamsGpu1 = vclMatrix(myStreamsCpu)
myStreamsGpu2 = createStreamsGpu(n=4, initial=12345)
range(as.matrix(myStreamsGpu1 - myStreamsGpu2))
## [1] 0 0
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{createStreamsGpu} also has the two arguments \code{n} and \code{initial}. \fct{createStreamsGpu} is similar to \code{.Random.seed} in \proglang{R}. By setting the RNG status with the \proglang{R} function \fct{set.seed}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \fct{createStreamsGpu} and \fct{createStreamsCpu}, with the \code{initial} argument, we make the sequence of streams repeatable.

\subsection{Using streams to generate uniform random numbers}
The streams created sequentially on a GPU are then used in work-items that executes in parallel on a GPU device. In \pkg{gpuRandom} each work-item takes one distinct stream to generate random numbers, so the number of streams should always be equal to the number of total work-items in use. The main part of the kernel (\proglang{OpenCL} functions for execution on the device) for generating uniform random numbers is shown in Listing \ref{lst:uniformkernel}. Kernels are written in the \proglang{OpenCL C} language, in which \code{__kernel} declares a function as a kernel, pointer kernel arguments must be declared with an address space qualifier, for example \code{__global} and \code{__local}. Here the pointers to \code{streams} and to the output matrix \code{out} on the global memory are passed to the kernel as arguments. \code{Nrow} and \code{Ncol} represent row and column number of the matrix \code{out} respectively. \code{NpadCol} is the internal number of columns of \code{out}, \code{NpadStreams} is the internal number of columns of the \code{streams} (these variables are defined in macros not shown here). Each stream's current state is copied to the private memory of each work-item by the function \fct{streamsToPrivate}. \code{index} represents the position of the work-item when mapping it to a one-dimensional data structure. The function \fct{clrngMrg31k3pNextState} generates an uniform random integer called \code{temp} between 1 and 2147483647,  the value of \code{temp} is then scaled to be in the interval $(0,1)$ by multiplying it by a constant \code{mrg31k3p\_NORM\_cl}, which is defined to be $1/2147483648$, if to generate double-precision random numbers. At the end of generating random numbers, streams are transferred back to global memory through the function \fct{streamsFromPrivate}.
\begin{framed}
\begin{lstlisting}[language=C,basicstyle=\small,label={lst:uniformkernel},caption={Using streams to produce random U(0,1) numbers}]
 __kernel void mrg31k3pMatrix(
    __global int* streams,
    __global double* out){

@@

    }//Dcol
 }//Drow
streamsFromPrivate(streams,g1,g2,startvalue);
}
\end{lstlisting}
\end{framed}
This kernel executes over a 2-dimensional NDRange. Figure \ref{fig:1} illustrates how this kernel executes in practice with a simple example. Suppose we have a $2 \times 4$ NDRange ((a) in Figure \ref{fig:1}) and want to create a $4 \times 6$ matrix ((b) in Figure \ref{fig:1}) of uniform random numbers. In each grid of (a) that represents a work-item, there are two rows of numbers, the number in the above row is the stream index (from 0 to 7), and a pair of numbers in the second row is the 2-dimensional global ID of the work-item. Each stream object is passed to each work-item. The number in a cell of the matrix corresponds to the stream index, it indicates which stream generates a random number for this matrix cell. For example, stream 0 which is allocated in work-item $(0,0)$ generates a random number in matrix $(0,0)$ in the first iteration, then moves on to matrix $(0,2)$ and generates a random number there in the second iteration, in the third iteration it generates a random number in matrix $(0,4)$ and so on until it goes out of the matrix range in the next iteration.
\begin{figure}[H]
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f1} }}
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f2} }}
    \caption{Using streams to produce a $4 \times 6$ matrix of random U(0,1) numbers. \label{fig:1}}
\end{figure}

Now we use the streams created in section \ref{createstreams} to generate two vectors of double-precision i.i.d. $U(0,1)$ random numbers with the function \fct{runif}. \fct{runif} can also be used to produce random integers in $[1, 2147483647]$. To view the generated random numbers, we need to convert them to R vectors or matrices, by doing this, the random numbers are moved from the global memory of the GPU device to the host.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
as.vector(gpuRandom::runif(n = 6, streams = myStreamsGpu1, Nglobal = c(1,
    nrow(myStreamsGpu1)), type = "double"))
## [1] 0.7353245 0.5180770 0.6142074 0.2319392 0.1100781 0.3619766
as.vector(gpuRandom::runif(n = 10, streams = myStreamsGpu2, Nglobal = c(1,
    nrow(myStreamsGpu1)), type = "double"))[-(1:6)]
## [1] 0.6487742 0.1112075 0.3661944 0.5018562
\end{verbatim}
\end{kframe}
\end{knitrout}
The arguments of the \fct{gpuRandom::runif} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams for random number generation.
  \item \code{Nglobal}: global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' or ``integer'' format of generated random numbers.
\end{itemize}
\fct{runif} returns the same sequence of random numbers because it uses the same streams \code{myStreamsGpu1} and \code{myStreamsGpu2} here. Reusing \code{myStreamsGpu1} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream advances by one position. %Below shows the \code{myStreamsGpu1} after it being used.

@@

\subsection{Plot the uniform random numbers (now deleted?)}
%We evaluate the distribution of 50000 GPU-generated double-precision uniform random numbers by making a density histogram  (the left panel in Figure \ref{fig0}), and plotting its empirical cumulative distribution and overlaying the theoretical cumulative distribution (the right panel in Figure \ref{fig0}). The two cumulative distribution lines seem to overlap completely, suggesting the data generated by \fct{runif} does come from a standard uniform distribution.

%\begin{figure}[H]
%\centering

%\caption{Histogram and empirical cumulative distribution of the GPU-generated random numbers.\label{fig0}}
%\end{figure}

\section{Some non-uniform random number generation}
\subsection{Normal random number generation}
We apply the Box-Muller transformation to $U(0,1)$ random numbers to generate standard normal random numbers. As shown in Algorithm \ref{algorithm1}, Box-Muller algorithm takes two independent, standard uniform random variables $U_1$ and $U_2$ and produces two independent, standard Gaussian random variables $X$ and $Y$, where $R$ and $\Theta$ are polar coordinate random variables. %The derivation basically uses transformation from Cartesian coordinates to polar coordinates to represent two independent, Normally distributed random variables. 
The Box-Muller algorithm turns out to be the best choice for Gaussian transform on GPU compared to other transform methods \citep{howes2007efficient}, because this algorithm has no branching or looping, which are the things GPU is not good at. %but GPU is good at the high computational load of sine and cosine functions in the algorithm.

@@

\begin{figure}[H]%
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f3} }}%
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f4} }}%
    \caption{Using streams to produce a $4 \times 6$ matrix of normal random numbers.}%
    \label{fig3}
\end{figure}

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{gpuRandom::rnorm}. The best performance we have seen using \fct{gpuRandom::rnorm} is around 120 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n = 512 * 128)
system.time(gpuRandom::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.142   0.003   0.145
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   6.689   0.830   7.513
\end{verbatim}
\end{kframe}
\end{knitrout}

\subsection{Exponential random number generation}
The Exponential random variates are produced by applying the inverse transform method on i.i.d. $U(0,1)$ random numbers. The random variable $X \sim \text{Exponential}(\lambda)$ has cumulative distribution function $F_{X}(x)=1-e^{-\lambda x}$ for $x \geq 0$ and $\lambda > 0$. The inverse of $F^{-1}_{X}$ is $F^{-1}_{x}(y)= -(1/ \lambda) \log (1-y)$, for $ 0 \leq y <1$. The kernel is not shown because it is mostly same as the kernel for uniform and normal random numbers, basically just by applying the inverse transform formula to the \code{clrngMrg31k3pNextState(g1, g2)} in the kernel. Developers can make use of our kernel for generating other types of random variables on GPU in \proglang{R}.

Below is the code that we produce a 2 by 2 matrix of exponential random number with expectation 1.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
r_matrix <- gpuRandom::rexp(c(2,2), rate=1, Nglobal=c(1,4), type="double")
as.matrix(r_matrix)
##           [,1]      [,2]
## [1,] 0.9260462 0.1431169
## [2,] 0.6722247 0.8337073
\end{verbatim}
\end{kframe}
\end{knitrout}
The arguments of the \fct{gpuRandom::rexp} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams. If missing, a stream with random initial state is supplied.
  \item \code{Nglobal}: global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
\end{itemize}

\section{An application of uniform RNG: Fisher's simulation}
The GPU-generated random numbers can be applied in suitable statistical simulations to accelerate the performance.
One application of GPU-generated random numbers in \pkg{gpuRandom} is Monte Carlo simulation for Fisher's exact test. Fisher’s exact test is applied for analyzing usually $2 \times 2$ contingency tables when one of the expected values in table is less than 5. Different from methods which rely on approximation, Fisher's exact test computes directly the probability of obtaining each possible combination of the data for the same row and column totals (marginal totals) as the observed table, and get the exact p-value by adding together all the probabilities of tables as extreme or more extreme than the one observed. However, when the observed table gets large in terms of sample size and table dimensions, the number of combinations of cell frequencies with the same marginal totals gets very large, \cite[][p. 23]{mehta2011ibm} shows a $5 \times 6$ observed table that has 1.6 billion possible tables. Calculating the exact P-values may lead to very long run-time and can sometimes exceed the memory limits of your computer. Hence, the option \code{simulate.p.value = TRUE} in \fct{stats::fisher.test} is provided, which enables computing p-values by Monte Carlo simulation for tables larger than 2 $\times$ 2.

@@

The arguments of the \fct{gpuRandom::fisher.sim} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{x}: a ``vclMatrix''.
  \item \code{N}: number of simulation runs.
  \item \code{streams}: streams.
  \item \code{type}: ``double'' or ``float'' of returned test statistics.
  \item \code{returnStatistics}: if TRUE, return all test statistics. 
  \item \code{Nglobal}: global index space. Default is set as $(64,16)$.
\end{itemize}
We show the advantage of \fct{gpuRandom::fisher.sim} by computing the p-values for two real data examples: one with a relatively big p-value and another with a very small p-value, and compare the run-time with using \fct{stats::fisher.test} for each of the data sets on two computers: one with a very good CPU and an ordinary GPU, the other is equipped with an excellent GPU and an ordinary CPU. The \proglang{R} outputs for testing on computer 2 is shown in the following.

@@

\subsection{Comparing run-time: Month data example}\label{fisher_month}

@@

month_gpu<-vclMatrix(month,type="integer")
result_month <- fisher.sim(month_gpu, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(256,64))
result_month$simNum
## [1] 1015808
result_month$counts
## [1] 410825
result_month$p.value
## [1] 0.4044323
\end{verbatim}
\end{kframe}
\end{knitrout}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(fisher.sim(month_gpu, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = c(256,64)))
##    user  system elapsed 
##   0.739   0.004   0.742
\end{verbatim}
\end{kframe}
\end{knitrout}
We requested one million simulations on GPU while the actual number of simulation is 1015808, and got 409524 cases whose test statitsic is below the observed threshold, so the p-value from \fct{gpuRandom::fisher.sim} for this table is about 0.40443, which is close enough to the p-value from \fct{stats::fisher.test}: 0.40448. \fct{gpuRandom::fisher.sim} takes about 0.69 seconds, \pkg{gpuRandom} accounts for 4.64\% of the elapsed time on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
stats::fisher.test(month,simulate.p.value = TRUE,B=1015808)$p.value
## [1] 0.4044816
system.time(stats::fisher.test(month,simulate.p.value = TRUE,B=1015808))
##    user  system elapsed 
##  15.006   0.009  15.002
\end{verbatim}
\end{kframe}
\end{knitrout}

\subsection{Comparing run-time: Week data example}\label{fisher_week}
